A mechanical Model of Rod-Shaped Bacteria

### 2025-06-19 paper/main.tex
HUGE update for paper; this could be an initial draft from where we start iterating

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage[hidelinks]{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsmath,amssymb,amsthm}
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{mathrsfs}
\usepackage{onimage}
\usepackage{graphicx}
\usepackage{doi}
\usepackage{acronym}
\usepackage{listings}
\usepackage{tikz}
\usepackage{tabularx}

@@

@@

%###################################################################################################
\begin{abstract}
    This paper develops the formal notions of cellular \dots
\end{abstract}

@@

@@

Bacteria exist in many physical shapes such as spheroidal, rod-shaped or spiral
\cite{Zapun2008,Young2006}.

To model the spatial mechanics of elongated bacteria \cite{Billaudeau2017}, we represent them as a
collection of auxiliary vertices $\{\vec{v}_i\}$ which are connected by springs in
ascending order.
Furthermore, we assume that the cells are flexible described by their rigidity.
A force $\vec{F}$ interacting between cellular agents determines the radius (thickness) of the
rods and an attractive component can model adhesion between cells.
\begin{table}
    \centering
    \def\arraystretch{1.3}
    \begin{tabularx}{\textwidth}{c l X}
        &\textbf{Aspect} & \textbf{Description}\\
        \toprule
        &\textbf{(C) Cellular}\\
        \midrule
        (1) & Rod-Shaped Mechanics &
            Rod-shaped bacteria are flexible rods which are able to freely move around (off-lattice
            approach) \cite{Takeuchi2005,Ursell2014,Amir2014_2}.\\
        (2) & Growth &
            Cells grow exponentially by inserting new material along the circular part of the rod
            \cite{Robert2014,Takeuchi2005}.\\
        (3) & Differentiation &
            \textit{B.subtilis} is known to differentiate into matrix-producing and
            surfactin-producing cells \cite{vanGestel2015,Lpez2010}.\\
        (4) & Division &
            The formation and continuation of van Gogh bundles is driven by cell division
            \cite{vanGestel2015}.
            Bacteria can form multilayers during their growth phase \cite{Duvernoy2018}.\\
        (5) & Variable Parameters &
            Parameters for individual cells are not fixed values but rather taken from a
            distribution \cite{Koutsoumanis2013}.\\
        &\textbf{(CC) Cell-Cell Interactions}\\
        \midrule
        (6) & Adhesion &
            Bacteria adhere to each other at longer distances and attach when in close contact
            \cite{Verwey1947,Trejo2013}.
            The interaction between rods can be polarized \cite{Duvernoy2018}.\\
        (7) & Friction &
            Friction between cells \cite{Grant2014} can be asymmetrical \cite{Doumic2020}.\\
        &\textbf{(DC) Domain-Cell Interactions}\\
        \midrule
        (8) & External Forces &
            Bacteria tend to stick to surfaces \cite{vanLoosdrecht1989}.
            The extracellular gel exerts a force onto the cells \cite{Grant2014}.\\
        (9) & Extracellular Reactions &
            Bacteria can take up nutrients or possible secrete/take up signalling molecules
            \cite{Li2025}.\\
        \bottomrule
    \end{tabularx}
    \label{table:simulation-aspects}
    \caption{TODO Insert caption}
\end{table}

% --------------------------------------------------------------------------------------------------
\subsection{Mechanics}

@@

In principle we can assign individual lengths $\{l_i\}$ and strengths $\{\gamma\}_i$ to each spring.
The internal force acting on vertex $\vec{v}_i$ can be divided into 2 contributions coming
from the 2 springs pulling on it.
In the case when $i=0,N_\text{vertices}$, this is reduced to only one internal component.
We denote with $\vec{c}_{i}$ the connection between two vertices

\begin{equation}
    \vec{c}_i = \vec{v}_{i}-\vec{v}_{i-1}
\end{equation}

and can write down the resulting force

\begin{align}
    \vec{F}_{i,\text{springs}} =
        &-\gamma_i\left(1 - \frac{l_i}{\left|\vec{c}_i\right|}\right)\vec{c}_i\\
        &+ \gamma_{i+1}\left(1 - \frac{l_{i+1}}{\left|\vec{c}_{i+1}\right|}\right)\vec{c}_{i+1}
    \label{force-springs}
\end{align}

In addition to springs between individual vertices $\vec{v}_i$, we assume that each angle at
a vertex between two other is subject to a curvature force.
Assuming that $\alpha_i$ is the angle between the connections and
$\vec{d}_i=\vec{c}_i/|\vec{c}_i|$ is the normalized connection,
we can write down the forces acting on vertices $\vec{v}_i,\vec{v}_{i-1},\vec{v}_{i+1}$

\begin{align}
    \vec{F}_{i,\text{curvature}} &= \eta_i\tan\left(\pi-\alpha_i\right)
        \frac{\vec{d}_i - \vec{d}_{i+1}}{|\vec{d}_i-\vec{d}_{i+1}|}\\
    \vec{F}_{i-1,\text{curvature}} &= -\frac{1}{2}\vec{F}_{i,\text{curvature}}\\
    \vec{F}_{i+1,\text{curvature}} &= -\frac{1}{2}\vec{F}_{i,\text{curvature}}
    \label{force-curvature}
\end{align}

where $\eta_i$ is the curvature at vertex $\vec{v}_i$.
We can see that the curvature force does not move the overall center of the cell in space.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]
        {../docs/source/_static/mechanics.png}
    \caption{Example of a polygonal representation of a rod-shaped bacterium.}
    \label{fig:mechanics-bacterium}
\end{figure}

The total force is the sum of external and interal forces.

\begin{equation}
    \vec{F}_{i,\text{total}} = \vec{F}_{i,\text{springs}}+ \vec{F}_{i,\text{curvature}} + \vec{F}_{i,\text{external}}
   \label{force-total}
\end{equation}

\begin{align}
    \partial_t^2 \vec{x} &= \partial\vec{x} + D\vec{\xi}\\
    \partial_t\vec{x} &= \vec{F}_\text{total}
    \label{equations-of-motion}
\end{align}

where $D$ is the diffusion constant and  $\vec{\xi}$ is the wiener process (compare with
brownian motion such as given by the
% Brownian3D <https://cellular-raza.com/docs/cellular_raza_building_blocks/struct.Brownian3D.html>_
% struct of cellular-raza <https://cellular-raza.com>.

% --------------------------------------------------------------------------------------------------
\subsection{Interaction}

When calculating forces acting between the cells, we can use a simplified model to circumvent the
numerically expensive integration over the complete length of the rod.
Given a vertex $\vec{v}_i$ on one cell, we calculate the closest point $\vec{p}$ on the polygonal
line given by the vertices $\{\vec{w}_j\}$ of the interacting cell.
Furthermore we determine the value $q\in[0,1]$ such that

\begin{equation}
    \vec{p} = (1-q)\vec{w}_j + q\vec{w}_{j+1}
   \label{connection}
\end{equation}

for some specific $j$.
The force is then calculated between the points $\vec{v}_i$ and $\vec{p}_i$ and acts on
the vertex $\vec{w}_i,\vec{w}_{i+1}$ with relative strength $(1-q)$ and $q$.

\begin{equation}
    \vec{F}_{i,\text{External}} = \vec{F}(\vec{v}_i,\vec{p})
    \label{force-external}
\end{equation}

% .. _ TODO MORSEPOTENTIAL

% .. _ TODO insert table with all parameters

@@

To simulate proliferation, we introduce a growth term for the spring lengths $l_i$

\begin{equation}
    \partial_t l_i = \mu
    \label{growth-ode}
\end{equation}

which will increase the length of the cell indefenitely unless we introduce a
% `division event <https://cellular-raza.com/internals/concepts/cell/cycle>`_.
We define a threshold (in our case double of the original length) for the total length of the
cell at which it divides.
To construct a new cell, we cannot simply copy the existing one twice, but we also need to adjust
internal parameters in the process.
The following actions need to be taken for the old and new agent.

% .. _ TODO

1. Assign a new growth rate (pick randomly from uniform distribution in $[0.8\mu_0,1.2\mu_0]$
   where $\mu_0$ is some fixed value)
2. Assign new positions
    1. Calculate new spring lengths
        $\tilde{l}_i = l_i\left(\frac{1}{2} - \frac{r}{\sum\limits_i l_i}\right)$
    2. Calculate middle of old cell
        $\vec{m} = \frac{1}{N_\text{vertices}}\sum\limits_i\vec{v}_i$
    3. Calculate positions of new vertices $\vec{w}_i$

\begin{align}
    q_i &= \frac{i}{N_\text{vertices}}\\
    \vec{w}_{i,\text{new},\pm} &= (1-q_i)\vec{m} + q_i(\vec{v}_{\pm\text{start}} - \vec{m})
\end{align}

%###################################################################################################
\section{Generating Masks and Images}
\label{section:generating-masks-and-images}

\begin{figure}[!h]
    \centering
    \begin{tikzonimage}[width=0.3\textwidth]
        {../docs/source/_static/09395645494836445480/raw_pv/000000400.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{a}};
    \end{tikzonimage}
    \begin{tikzonimage}[width=0.3\textwidth]
        {../docs/source/_static/09395645494836445480/images/000000400.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{b}};
    \end{tikzonimage}
    \begin{tikzonimage}[width=0.3\textwidth]
        {../docs/source/_static/09395645494836445480/masks/000000400.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{c}};
    \end{tikzonimage}
    \caption{Snapshots at \dots}
    \label{fig:progression-image-generation}
\end{figure}

%###################################################################################################
\section{Extract Positions from Masks}
\label{section:fitting-and-comparing-masks}

\begin{figure}
    \centering
    \begin{tikzonimage}[width=0.32\textwidth]
        {../docs/source/_static/fitting-methods/algorithm/image001042.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{a}};
    \end{tikzonimage}
    \begin{tikzonimage}[width=0.32\textwidth]
        {../docs/source/_static/fitting-methods/algorithm/mask-zoom.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{b}};
    \end{tikzonimage}
    \begin{tikzonimage}[width=0.32\textwidth]
        {../docs/source/_static/fitting-methods/algorithm/mask-zoom.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{c}};
    \end{tikzonimage}
    \caption{}
    \label{fig:position-extraction-algorithm}
\end{figure}

\begin{figure}
    \centering
    \begin{tikzonimage}[width=0.32\textwidth]
        {../docs/source/_static/fitting-methods/extract_positions-004000.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{a}};
    \end{tikzonimage}
    \begin{tikzonimage}[width=0.32\textwidth]
        {../docs/source/_static/fitting-methods/extract_positions-006000.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{b}};
    \end{tikzonimage}
    \begin{tikzonimage}[width=0.32\textwidth]
        {../docs/source/_static/fitting-methods/extract_positions-008000.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{c}};
    \end{tikzonimage}
    \caption{Progression of generated masks with estimated and exact positions.}
    \label{fig:position-extraction}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]
        {../docs/source/_static/fitting-methods/displacement-calculations.png}
    \caption{Comparison of masks across generations. Compare estimated with exact rod lengths.}
    \label{fig:benchmarking-fitting-algorithm}
\end{figure}

%###################################################################################################
\section{Estimating Parameters}
\label{section:parameter-estimation}

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage[hidelinks]{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsmath,amssymb,amsthm}
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{float}
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{mathrsfs}
\usepackage{onimage}
\usepackage{graphicx}
\usepackage{doi}
\usepackage{acronym}
\usepackage{listings}
\usepackage{tikz}
\usepackage{tabularx}
\usepackage{siunitx}
\usepackage{cite}
\usepackage{acronym}

@@

\newacro{abm}[ABM]{Agent-Based Model}
\newacro{ode}[ODE]{Ordinary Differential Equation}
\newacro{pbr}[PBR]{Physically-Based Rendering}

@@

%###################################################################################################
\begin{abstract}
    This paper establishes the basis for a novel individual-based mechanical model of rod-shaped
    bacteria.
    It describes methods of how to obtain data from microscopic images such that they can be used
    for parameter estimation purposes.
    Furthermore, we show initial results for parameter estimates and an application of the model to
    the formation of multilayers within bacterial colonies.
\end{abstract}

@@

Bacteria exist in many physical shapes such as spheroidal, rod-shaped or spiral
\cite{Zapun2008,Young2006}.
The last decades have fundamentally challenged and changed our understanding of how bacteria retain
their shape.
Rod-shaped bacteria can grow by extending their circular part, or by inserting new material at the
tip of the rod.
Species such as E.coli and B.subtilis~\cite{Errington2020} fall under the first category while
S.pombe represents the latter.
The material is inserted in small bursts on the nanoscale in forms of patches, bands or
hoops~\cite{DePedro2003}.
Despite their differences in wall-size, B.subtilis and E.coli follow growth rules which are
comparable with respect to the extension of the rod~\cite{Chang2014}.

\begin{itemize}
    \item This publication contains 5 things:
    \begin{enumerate}
        \item Initial (incomplete) model of rod-shaped bacteria
        \item Methods of how to obtain data from microscopic images and perform parameter
            estimations; $\rightarrow$ also possibly across multiple generations
        \item Showcase the usefulness of the methods and the model by applying it to parameter
            estimation via microscopic images and multilayer formation
        \item Generate 3D Images (masks)
        \item A python package which can do all of that and can be reused and extended
    \end{enumerate}
    \item Make clear that this is mainly a proof of the methods
    \item For complete model, more work has to be done
    \begin{itemize}
        \item More aspects (see Review; put in discussion)
        \item Note that we only care about \textit{E.Coli} and \textit{B.Subtilits}! This is
            important.
    \end{itemize}
\end{itemize}

% TODO write more text here
% Use some parts of the proposal text

@@

In order to construct a mechanical model of rod-shaped bacteria, we have to take care of a number of
different aspects.
The mechanical behaviour of rod-shaped bacteria is governed by a plethora of effects.
Although we set out to construct a model which is able to describe as many of these effects as
possible, we will fall short in this paper and restrict ourselves to the aspects presented
in~\ref{table:simulation-aspects}.
We will discuss, how this model can be extended in order to capture more effects and how we can
validate our approach given the existing literaturesection:supplement-more-simulation-aspects
(see section~\ref{section:discussion}, and~\ref{section:supplement-more-simulation-aspects}).

\paragraph{(C) Cellular}
The key component of this model consists of the cellular representation.

It was long believed that bacteria lack cytoskeletal filaments but with the end of the last century,
it was discovered that the MreB protein fills this role since it “[..] forms an actin-like
cytoskeleton in bacteria [..]”~\cite{Erickson2001}.
It can polymerize and form filaments which behave similar to actin microfilaments~\cite{Dersch2020}.
The effect of MreB on the shape of cells during growth was shown in more detail
by~\cite{Ursell2014}.
They found that MreB localizes specifically at regions of negative curvature and thus introduces
heterogeneous cell growth which results in a bending force during the growth of cells.
It also means that the cells are able to sense the curvature of the cell wall through cytoskeleteal
MreB filaments during growth.
\cite{Bratton2018} showed that the trans-membrane protein RodZ can modulate MreB density and
facilitates local MreB assembly.
This feedback mechanism will result in straight rod-shaped bacteria if not interfered from the
external environment.
These results were also confirmed by the studies of~\cite{Wang2010}.
In their experiments, a single bacterium was fixed at the bottom pole while an optically trapped
bead was placed on the upper pole.
They could now exert a force on the free part and thus study the response of the cell under
different conditions.
When treating cells with A22, an antibiotic which disassembles
MreB~\cite{IWAI2002,Gitai2005,Karczmarek2007,Bean2009}, a higher displacement with identical force
could be measured, thus indicating that MreB increases mechanical rigidity of the cell.

While some rod-shaped bacteria grow by inserting new material at the ends, \textit{E.Coli} and
\textit{B.Subtilis} utilize a method where new material is inserted along the cylindrical part of
the rod.
The details of the division event (i.e. Z-Ring formation) are very intricate and are heavily
studied as a subject on their own (TODO citation).
On a higher level view, the rod is split apart at a point close to the center of the cell, thus
creating two new daughter cells.
We know that the best predictor for when the division event occurs, is the length of the rod as
opposed to the age of bacterium~\cite{Robert2014}.
Furthermore, only the growth of the whole colony but also the individual growth of the bacteria is
exponential~\cite{Amir2014,Takeuchi2005}.

\paragraph{(CC) Cell-Cell Interactions \& (DC) Domain-Cell Interactions}
A colony of bacteria is known to form biofilms (Dunne 2002). This strategy is is beneficial for the
survival of the collective as it allows it to harness growth factors and migrate when necessary and
also plays an important role in bacterial infections in the biomedical sector~\cite{Ong1999}.
The formation of biofilms is only possible due to the attractive nature of bacterial interactions
towards each other and surfaces (Berne et al. 2018). To analyze bacterial adhesion to surfaces,
\cite{vanLoosdrecht1989} investigated the interactions of various bacteria with negatively charged
polystyrene.
They found that the interaction is characterized by a low Gibbs energy of $(2 - 3)kT$ per cell
which can be described by a secondary minimum of the DLVO-theory~\cite{Verwey1947,Derjaguin1993}
which is relevant for distances $\geq\SI{1}{\nano\meter}$.
At these distances, the DLVO potential acts as a generic adhesive potential.
Particles which stay in the secondary minimum are held together but even weak interactions are
enough to redisperse them which means that this state is reversible.
When approaching closer distances, more intricate mechanisms come into play in which the
heterogeneity of the materials in question needs to be considered.
\cite{Hori2010} described bacterial adhesion as a “two-phase process including an initial,
instantaneous and reversible physical phase (phase one) and a time-dependent and irreversible
molecular and cellular phase (phase two)”.

All these considerations become even more intricate when not considering interactions of bacteria
with surfaces but bacteria with each other.
Gram-negative bacteria (eg. E.Coli) express outer membrane proteins (adhesins) which allow them to
attach to various hosts (Beachey 1981, Vaca et al. 2019).
In addition, the flagella can play an important role in the direct adhesion between
cells~\cite{Haiko2013}.
Thus it is not correct to blindly apply the results for adhesion to surfaces to the interactions
between bacteria.

\begin{table}
    \centering
    \def\arraystretch{1.3}
    \begin{tabularx}{\textwidth}{c l X}
        &\textbf{Aspect} & \textbf{Description}\\
        \toprule
        &\textbf{(C) Cellular}\\
        \midrule
        (1) & Rod-Shaped Mechanics &
            Rod-shaped bacteria are flexible rods which are able to freely move around (off-lattice
            approach) \cite{Takeuchi2005,Ursell2014,Amir2014_2}.\\
        (2) & Growth &
            Cells grow exponentially by inserting new material along the circular part of the rod
            \cite{Robert2014,Takeuchi2005}.\\
        % (3) & Differentiation &
        %     \textit{B.subtilis} is known to differentiate into matrix-producing and
        %     surfactin-producing cells \cite{vanGestel2015,Lpez2010}.\\
        (3) & Division &
            Rod-Shaped bacteria divide in the middle of the rod, thus forming two new
            agents.
            Bacteria can form multilayers during their growth phase \cite{Duvernoy2018}.\\
        (4) & Variable Parameters &
            Parameters for individual cells are not fixed values but rather taken from a
            distribution \cite{Koutsoumanis2013}.\\
        &\textbf{(CC) Cell-Cell Interactions}\\
        \midrule
        (5) & Adhesion &
            Bacteria adhere to each other at longer distances and attach when in close contact
            \cite{Verwey1947,Trejo2013}.\\
            % The interaction between rods can be polarized \cite{Duvernoy2018}.\\
        % (7) & Friction &
        %     Friction between cells \cite{Grant2014} can be asymmetrical \cite{Doumic2020}.\\
        &\textbf{(DC) Domain-Cell Interactions}\\
        \midrule
        (6) & External Forces &
            Bacteria which rub against surfaces tend to stick to them after some
            time~\cite{vanLoosdrecht1989}.
            The extracellular gel which surrounds the bacteria exerts a force onto the cells
            \cite{Grant2014}.\\
        % (9) & Extracellular Reactions &
        %     Bacteria can take up nutrients or possible secrete/take up signalling molecules
        %     \cite{Li2025}.\\
        \bottomrule
    \end{tabularx}
    \label{table:simulation-aspects}
    \caption{TODO Insert caption}
\end{table}

To model the spatial mechanics of elongated bacteria \cite{Billaudeau2017}, we represent them as a
collection of auxiliary vertices $\{\vec{x}_i\}$ which are connected by springs in
ascending order.
Furthermore, we assume that the cells are flexible described by their rigidity.
A force $\vec{F}$ interacting between cellular agents determines the radius (thickness) of the
rods and an attractive component can model adhesion between cells.

% --------------------------------------------------------------------------------------------------
\subsection{Mechanics}
\label{subsection:mechanical-model-mechanics}

We use principles from discrete differential geometry~\cite{bobenko2008discrete} to describe the
stiffness and flexibility of the bacterial rod.
As explained previously, MReB will counteract the bending of the rod, resulting in a stiffening
force.
In order to construct a model which can be parametrized and by method of Occam's Razor, we chose to
model the bacterium as a bending beam, thus utilizing the simplest solution which can still produce
the desired effects.
Such a beam can be represented by a collection of vertices $\vec{x}_i$ which are connected by a
springs of lengths $l_i$.
We use the notation $\vec{c}_i:=\vec{x}_i-\vec{x}_{i-1}$ with which the resulting force acting
between two vertices if given by

\begin{align}
    \vec{F}_{i,\text{springs}} =
        &-\gamma\left(1 - \frac{l}{\left|\vec{c}_i\right|}\right)
        \vec{c}_i\\
        &+ \gamma\left(1 - \frac{l}{\left|\vec{c}_{i+1}\right|}\right)
        \vec{c}_{i+1}.
\end{align}

In addition to springs between individual vertices $\vec{x}_i$, we assume that each angle at a
vertex between edges is subject to a force indiced by curvature.
% We assume that we can model the mechanical properties of the bacterium as an elastic rod.
Given the angle $\alpha_i = \sphericalangle(\vec{c}_{i-1},\vec{c}_i)$ between adjacent edges,
the bending force is proportional to the curvature $\kappa_i$ at each vertex
$\kappa_i = 2\tan(\alpha_i/2)$.
The resulting force acts along the angle bisector which can be calculated from the edge vectors.
Figure~\ref{fig:mechanics-bacterium} shows an elongated rod represented by $6$ vertices.
The forces acting on vertices $\vec{x}_i,\vec{x}_{i-1},\vec{x}_{i+1}$ are given by

\begin{align}
    \vec{F}_{i,\text{curvature}} &= \eta\kappa_i
        \frac{\vec{c}_i - \vec{c}_{i+1}}{|\vec{c}_i-\vec{c}_{i+1}|}\\
    \vec{F}_{i-1,\text{curvature}} &= -\frac{1}{2}\vec{F}_{i,\text{curvature}}\\
    \vec{F}_{i+1,\text{curvature}} &= -\frac{1}{2}\vec{F}_{i,\text{curvature}}
\end{align}

where $\eta_i$ is the rigidity at vertex $\vec{x}_i$ (see Figure~\ref{fig:mechanics-bacterium}).
We can see that the curvature force does not move the overall center of the rod in space.
The total force is the sum of external and interal forces.

\begin{equation}
    \vec{F}_{i,\text{total}} = \vec{F}_{i,\text{springs}}+ \vec{F}_{i,\text{curvature}}
        + \vec{F}_{i,\text{external}}
\end{equation}

\begin{align}
    \partial_t^2 \vec{x}_i &= \partial_t\vec{x}_i + \sqrt{2D}\vec{\xi}_i\\
    \partial_t\vec{x}_i &= \vec{F}_{i,\text{total}} - \lambda \partial_t\vec{x}_i
\end{align}

where $D$ is the diffusion constant and $\vec{\xi}$ is the wiener process.
Together they determine the stochastic motion of the individual vertices which can arise from
recurring collisions with small particles of the surrounding medium or other minor fluctuations.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../docs/source/_static/mechanics.png}
    \caption{
        Example of an elonged bacterial rod represented by a collection of vertices which are
        connected by springs.
        The angles $\alpha_i$ have been exaggerated such that they can be identified more easily.
        A bacterial agent which is not interacting with any external objects will tend to the
        equilibrium state of a straight rod with $0$ curvature.
    }
    \label{fig:mechanics-bacterium}
\end{figure}

% --------------------------------------------------------------------------------------------------
\subsection{Interactions}

When calculating forces acting between the cells, we can use a simplified model to circumvent the
numerically expensive integration over the complete length of the rod.
Given a vertex $\vec{x}_i$ on one cell, we calculate the closest point $\vec{p}$ on the polygonal
line given by the vertices $\{\vec{w}_j\}$ of the interacting cell.
Furthermore we determine the value $q\in[0,1]$ such that $\vec{p} = (1-q)\vec{w}_j + q\vec{w}_{j+1}$
for some specific $j$.
The force is then calculated between the points $\vec{x}_i$ and $\vec{p}_i$ and acts on
the vertex $\vec{w}_i,\vec{w}_{i+1}$ with relative strength $(1-q)$ and $q$.

\begin{equation}
    \vec{F}_{i,\text{External}} = \vec{F}(\vec{x}_i,\vec{p})
    \label{eq:cell-force-external}
\end{equation}

The specific choice of interaction potential plays of course a crucial role for the dynamics of the
agents.
But we cannot expect to simply know the shape and parameters of this potential.
Thus, we would like to experiment with various interaction potentials in order to quantify their
ability to produce accurate results.
We utilize two distinct potentials that are both able to capture the essence of our interactions by
providing a short-range repulsive part and a long-ranged attractive component.

\paragraph{Morse Potential}
The morse potential~\cite{Morse1929} is a common example of numerically inexpensive interaction
potential.
It was originally designed to describe the interaction of diatomic molecules but has since found use
in other areas~\cite{Breitwieser2021}.
It contains a repulsive and attractive part and is bound from above which makes it particularly
suitable for numerical simulations since unbound potentials can lead to large numerical deviations
thus compromising the prediction of the model.
The potential is of the form
\begin{equation}
    V(r) = V_0\left(1-e^{-\lambda(r-(R_1+R_2))}\right)^2
\end{equation}
and has 3 parameters.
The overall potential strength $V_0$, the radii of the particles $R_1,R_2$ and the potential
stiffness $\lambda$ which controls the rate of change as one travels along the radial axis.
We employ a spatial cutoff which sets the interaction strength to $0$ when exceeding a distance
$r\geq\xi$.

\paragraph{Mie Potential}
The Mie~\cite{Mie1903} potential is the generalization of the well-known
Lennard-Jones~\cite{Jones1924} potential.
It was also designed to describe the interactions of molecules but due to the flexible nature of its
parameters, can be easily adapted to other situations.
It is able to describe a wider range of shapes compared to the Morse potential but is also stiffer
and not bound from above.
In our simulation, we enforce an upper bound to guarantee numerical regularity.
It has 4 parameters and is given by
\begin{align}
    U(r) &= C\epsilon\left[ \left(\frac{\sigma}{r}\right)^n -
        \left(\frac{\sigma}{r}\right)^m\right]\\
    C &= \frac{n}{n-m}\left(\frac{n}{m}\right)^{\frac{n}{n-m}}\\
    V(r) &= \min(U(r), \beta)\theta(r-\zeta)
\end{align}
where $C$ is a constant that depends on the two exponents $n,m$ that describe the overall shape of
the potential.
The strength is given by $\epsilon$ and the radius is encoded in $\sigma =
(m/n)^{1/(n-m)}(R_1+R_2)$ where $R_1,R_2$ are the radii of the two interacting particles.

@@

To simulate proliferation, we introduce a growth term for the spring lengths $l_i$.
In most of our simulations, we choose to use identical lengths for all segments (i.e. $l_i=l$).
However, in order to generalize our approach we assume that individual lengths of segments could
vary.
We reduce the growth rate with increasing number of neighbors $N$.
This behaviour can be toggled on or off thus removing the additional term in brackets of
equation~\ref{eq:growth-rod-lengths}.

\begin{equation}
    \partial_t l_i = \mu\left(1-\frac{\text{min}(n,N)}{N}\right)l_i
    \label{eq:growth-rod-lengths}
\end{equation}

Note that the growth rate is proportional to the rod length $l$ which leads to exponential growth.
This process will increase the length of the cell indefenitely unless we impose a condition for the
division event.
We define a fixed length threshold for the total length of the cell at which it divides.
To construct a new cell, we cannot simply copy the existing one twice, but we also need to adjust
internal parameters in the process.
The following actions need to be taken for the old and new agent.
We first calculate the new spring lengths of the two resulting rods.
\begin{equation}
    \tilde{l} = l_i\left(\frac{1}{2} - \frac{r}{\sum\limits_i l_i}\right)
    \label{eq:growth-rod-new-lengths}
\end{equation}
Afterwards, we start at one endpoint $\vec{v}_0$ and travel along the polygon defined by the
bacterial rod.
We stop when we have reached a length that matches the length of the new spring length $\tilde{l}$.
From here, we continue and iterate this procedure until we have found all new vertices $\vec{w}_i$.
Equation~\ref{eq:growth-rod-new-lengths} ensures that that the tips of the two new rods are touching
right after the division event.

% --------------------------------------------------------------------------------------------------
\subsection{External Forces}
TODO
\begin{itemize}
    \item Friction force against surface. perpendicular to motion; proportional to velocity.
    \item gel pressure coming from experiment (simply downwards force $F_z$)
\end{itemize}

% --------------------------------------------------------------------------------------------------
\subsection{Numerical Implementation}

We implemented the described model in \texttt{cellular\_raza}~\cite{Pleyer2025}.
Doing so has many benefits.
Not only, was a more simpler model already available which could be adapted to our needs.
But furthermore, due to the nature of \texttt{cellular\_raza} in that it does not assume any
particular cellular representation, we are free to change the underlying model at any time, even
making fundamental changes.
It is an \ac{abm} which is able to natively track cells through the course of a numerical simulation
and also keeps track of parent/daughter relationships.
This allows us to to experiment with different model implementations and still retain performance.
By leveraging the strong safety guarantees provided by the Rust programming language, we eliminate a
whole class of problems from the start.
% TODO CITATION
In addition, we are using the \texttt{pyo3} and \texttt{maturin} crates which allow us to easily
generate python bindings and thus interoperate between the two ecosystems.
While this particular model is more complex due to the nature of non-trivial (asymmetric)
interactions, we were still able to properly utilize the speedup from executing simulations in
parallalel using the built-in parallelization methods of \texttt{cellular\_raza} (see
Supplementary Material, Figure~\ref{fig:performance}).

%###################################################################################################
\section{Generating Masks and Images}
\label{section:generating-masks-and-images}
\begin{itemize}
    \item use pyvista (VTK) for Visualization
    \item always do 3D render. This way, we can have overalp between agents and produce more
        realistic masks.
    \item Assign unique color values for idividual agents.
        We have a 1:1 mapping between agent identifier and color! (for each simulation different
        though)
    \item use parallel projection; this should be most close to what microscopes measure due to the
        relatively large distance between object and lens.
    \item use some noise to smear out the rods; this is already similar to microscopic images; use
        more like this and deal with real-world defects to make it more realistic (future, not now)
    \item Figure~\ref{fig:progression-image-generation} shows steps in generating image.
    \item Reflectivity, placement of light etc. can already introduce some heterogeneity in the
        visual of the rod.
    \item Plan to implement noise on the mesh itself; make it have rough surface (like real rods as
        seen under electron microscopy)
\end{itemize}

\paragraph{Algorithm}
\begin{enumerate}
    \item Get position of cell
    \item Insert spheres at each vertex
    \item Insert cylinder along edges
    \item Combine individual objects into one big mesh
    \item Use \ac{pbr}
\end{enumerate}

\begin{figure}[H]
    \centering
    \begin{tikzonimage}[width=0.3\textwidth]
        {../docs/source/_static/09395645494836445480/raw_pv/000000400.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{A}};
    \end{tikzonimage}
    \begin{tikzonimage}[width=0.3\textwidth]
        {../docs/source/_static/09395645494836445480/images/000000400.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{B}};
    \end{tikzonimage}
    \begin{tikzonimage}[width=0.3\textwidth]
        {../docs/source/_static/09395645494836445480/masks/000000400.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{C}};
    \end{tikzonimage}
    \caption{
        (A) shows the raw result from combining sphere and cylinder meshes to obtain the shape of a
        bacterial rod.
        This render contains lighting coming from 2 light sources which thus produces a small glow
        on the upper part of the meshes.
        Subfigure (B) takes this image and applies noise to it in order to make it look more
        like an actual microscopic image.
        (C) The render is projected along the z-axis and flat numerical values are assigned to the
        bacteria.
        The resuling render can have overlaps in which case the color of the rod is displayed that
        is higher along the z-direction.
    }
    \label{fig:progression-image-generation}
\end{figure}

%###################################################################################################
\section{Parameter Estimation}

In order to effectively estimate the parameters of our model, we need to set up a cost function
which can be minimized.
In doing so, we hope to identify all parameters, given relevant experimental data.
We use microscopic images and construct a novel algorithm that can extract the position of the agent
from a provided cell-mask.
These cell-masks are easily obtainable by either using existing segmentation tools or manually
annottating images.
Following this, we can define a cost-function and an optimization scheme to minimize said function.

% --------------------------------------------------------------------------------------------------
\subsection{Extract Positions from Masks}
\label{section:fitting-and-comparing-masks}
